---
title: "sql-workshop"
author: "Yilin Lu"
output: html_document
---

Before you follow the directions below, please take a screenshot of your AWS console showing the running database and upload it to your repo.

## Connect to AWS MySQL Database
```{r}
#install.packages("DBI", "RMySQL")

library(DBI)
library(RMySQL)

db_user <- 'admin'
db_password <- 'testsql!'
db_name <- 'oudb'
db_host <- 'PASTE YOUR ENDPOINT HERE'
db_port <- 3306

mydb <- dbConnect(MySQL(), user = db_user, password = db_password, dbname = db_name, host = db_host, port = db_port)

summary(mydb)
```

## Load OU Data
```{r}
#Student demographic data
studentInfo <- read.csv("studentInfo.csv", header = TRUE)
#Student assessment data
studentAssessment <- read.csv("studentAssessment.csv", header = TRUE)
#Course data
courses <- read.csv("courses.csv", header = TRUE)
studentRegistration <- read.csv("studentRegistration.csv", header = TRUE)
```

## Write data to the DB using the DBI package
```{r}
#List the tables in the DB - should be zero
dbListTables(mydb)

#Write a new table to the DB
dbWriteTable(mydb, "studentInfo", studentInfo)
dbWriteTable(mydb, "studentAssessment", studentAssessment)
dbWriteTable(mydb, "courses", courses)
dbWriteTable(mydb, "studentRegistration", studentRegistration)

#List tables to see that table was added
dbListTables(mydb)

#Read a particular table
dbReadTable(mydb, 'studentInfo')

#EXERCISE 1
#Make two toy data sets with at least three variables and at least 30 rows each in them. Have a mix of numeric and character variables. Transfer these dataframes to your SQL database using the DBI commands. Name the tables whatever you like.

# 1) Connect to AWS MySQL database
library(DBI)
library(RMySQL)

db_user <- 'admin'
db_password <- 'testsql!'
db_name <- 'musicdb'
db_host <- 'database-project1.cwiivhq5gfpy.us-east-2.rds.amazonaws.com'
db_port <- 3306

db1 <- dbConnect(MySQL(), user = db_user, password = db_password, dbname = db_name, host = db_host, port = db_port)

summary(db1)

# 2) Load music data 
albums = read.csv("albums.csv", header = TRUE)
artists = read.csv("artists.csv", header = TRUE)

# 3) Write data to the db1
dbWriteTable(db1, "albums", albums)
dbWriteTable(db1, "artists", artists)
dbListTables(db1)

```

## Getting into SQL - READING
```{r}
#Query a portion of the database (always returns dataframe)
dbGetQuery(mydb, "SELECT * FROM studentInfo LIMIT 10;")

dbGetQuery(mydb, "SELECT * FROM studentInfo ORDER BY id_student LIMIT 10;")

dbGetQuery(mydb, "SELECT id_student, gender FROM studentInfo ORDER BY id_student DESC LIMIT 10;") #Order listed will be reflected in order in table

dbGetQuery(mydb, "SELECT id_student AS 'Student ID', gender FROM studentInfo LIMIT 10;") #SQL Standard says quotes for literal strings and double quotes for everything else but that conflicts with R

#Count the number of rows
dbGetQuery(mydb, "SELECT COUNT(*) FROM studentAssessment;")

#Using a WHERE statement on all columns
dbGetQuery(mydb, "SELECT COUNT(*) FROM studentAssessment WHERE score > 50;")

#Using a WHERE statement on a single column (will not include missing data)
dbGetQuery(mydb, "SELECT COUNT(score) FROM studentAssessment WHERE score > 50;")

#Using an AND statement
dbGetQuery(mydb, "SELECT COUNT(*) FROM studentAssessment WHERE score > 50 AND id_assessment = '1752';")

#EXERCISE 2
#Read one of your toy data tables, make sure the output is ordered in descending order, you rename one of the variables and the output is limited to the first 20 rows.

#Read the other table according to a condition of one of the variables.

# 1) Read album data table 
dbGetQuery(db1, "SELECT artist_id, album_title, genre, year_of_pub AS year FROM albums ORDER BY artist_id DESC LIMIT 20;")

# 2) Read artists data table 
dbGetQuery(db1, "SELECT * FROM artists ORDER BY zip_code LIMIT 20")

```

## Getting into SQL - UPDATING
```{r}
#Count rows
dbGetQuery(mydb, "SELECT COUNT(*) FROM studentAssessment;")

#Add a row
dbGetQuery(mydb, "INSERT INTO studentAssessment (id_assessment, id_student, date_submitted, is_banked, score) VALUES ('00001', '1', '20', '0', '50');")

#Count rows again
dbGetQuery(mydb, "SELECT COUNT(*) FROM studentAssessment;")

#View inserted row
dbGetQuery(mydb, "SELECT * FROM studentAssessment ORDER BY id_student LIMIT 10;")

#Add a row with missing values
dbGetQuery(mydb, "INSERT INTO studentAssessment (id_assessment, id_student, date_submitted) VALUES ('00001', '1', '20');")

#View inserted row
dbGetQuery(mydb, "SELECT * FROM studentAssessment ORDER BY id_student LIMIT 10;")

#Update a row
dbGetQuery(mydb, "UPDATE studentAssessment SET score = '20' WHERE id_student = 1;")

dbGetQuery(mydb, "SELECT id_student, score FROM studentAssessment ORDER BY id_student LIMIT 10;")

#Update a row with NULL
dbGetQuery(mydb, "UPDATE studentAssessment SET score = 'NULL' WHERE id_student = 6516;")

#Delete a row (destructive)
dbGetQuery(mydb, "DELETE FROM studentAssessment WHERE id_student = 1;")

dbGetQuery(mydb, "SELECT * FROM studentAssessment ORDER BY id_student LIMIT 10;")

#EXERCISE 3
#Insert a new row in one of your toy data tables leaving one variable empty. Change one value in your other table. Display your new tables. Delete the row you edited and the row you inserted. 

# For artists dataset 
# 1) Inset a new row, leaving zip_code variable empty
dbGetQuery(db1, "INSERT INTO artists (row_names, artist_name, role, year_of_birth, country) VALUES ('1', 'An example', 'rapper', '2021', 'China');")

# 2) Display new artists dataset with the new row
dbGetQuery(db1, "SELECT * FROM artists ORDER BY year_of_birth DESC LIMIT 10")

# 3) Delete the new row and display the dataset 
dbGetQuery(db1, "DELETE FROM artists WHERE year_of_birth = 2021;")
dbGetQuery(db1, "SELECT * FROM artists ORDER BY year_of_birth DESC LIMIT 10;")

# For albums dataset
# 1) Change one value in albums dataset
dbGetQuery(db1, "UPDATE albums SET num_of_tracks = '0' WHERE artist_id = 1;")

# 2) Display new albums dataset with the new value 
dbGetQuery(db1, "SELECT * FROM albums ORDER BY artist_id LIMIT 10;")

# 3) Delete the edited row and display the dataset 
dbGetQuery(db1, "DELETE FROM albums WHERE artist_id = 1;")
dbGetQuery(db1, "SELECT * FROM albums ORDER BY artist_id LIMIT 10;")

```

## Add/Deleting Table
```{r}
#Creating a new table in SQL
dbGetQuery(mydb,"CREATE TABLE test (
  score INTEGER, 
  student TEXT
  );")

dbListTables(mydb)

#Inserting data into the table
dbGetQuery(mydb, "INSERT INTO test VALUES ( 10, 'Amy' );")
dbGetQuery(mydb, "INSERT INTO test VALUES ( 11, 'Jen' );")
dbGetQuery(mydb, "INSERT INTO test VALUES ( 9, 'Frank' );")

dbGetQuery(mydb, "SELECT * FROM test;")

#Inserting a NULL row
dbGetQuery(mydb, "INSERT INTO test DEFAULT VALUES;") #Will not work use instead:

dbGetQuery(mydb,"INSERT INTO test (score, student) SELECT score, id_student FROM studentAssessment;")

#Delete a table
dbGetQuery(mydb, "DROP TABLE test;")

dbGetQuery(mydb, "SELECT * FROM test;") #This should produce an error since your table no longer exists

#Delete a table if it exists
dbGetQuery(mydb, "DROP TABLE IF EXISTS test;") #No error since it is only if it exists

#EXERCISE 4
#Create a table that is exactly the same as your first toy data table but this time use SQL commands. Display your new table. Then delete the original table. 

# 1) Create a new artists table 
dbGetQuery(db1, "CREATE TABLE artists_v1 (row_names INTEGER, artist_name TEXT, role TEXT, year_of_birth INTEGER, country TEXT, zip_code TEXT);")

dbListTables(db1)

# 2) Insert artists data into artists_v1 table 
dbGetQuery(db1, "INSERT INTO artists_v1 (row_names, artist_name, role, year_of_birth, country, zip_code) SELECT row_names, artist_name, role, year_of_birth, country, zip_code FROM artists;")

# 3) Display artists_v1 table with limit of 10 rows and count the number of rows  
dbGetQuery(db1, "SELECT * FROM artists_v1 LIMIT 10;")
dbGetQuery(db1, "SELECT COUNT(*) FROM artists_v1;")
# Conclusion: the displayed table and counted row of 50,000 with 5 variables from artists_v1 table are exactly the same as the original artists table

# 4) Delete the original artists table 
dbGetQuery(db1, "DROP TABLE artists;")
dbListTables(db1)

```

# NULL Value
```{r}
#NULL is a state (similar to R), represents the lack of a value. But is not compatible with R backend so this code doesn't work as part of dbGetQuery()

#This doesn't work because NULL is not a value
SELECT * FROM test WHERE score = NULL;

#Instead use
SELECT * FROM test WHERE score is NULL;

```

# Constraints
```{r}
#Create table where student column *cannot* be NULL
dbGetQuery(mydb,"CREATE TABLE test2 (
  score INTEGER, 
  student TEXT NOT NULL
  );")

dbGetQuery(mydb, "DROP TABLE IF EXISTS test2;")

dbGetQuery(mydb,"CREATE TABLE test2 (
   score INTEGER DEFAULT 0, 
   student TEXT
   );")

dbGetQuery(mydb,"INSERT INTO test2 (score, student) VALUES ('1', 'A');")
dbGetQuery(mydb,"INSERT INTO test2 (student) VALUES ('B');")

dbGetQuery(mydb, "SELECT * FROM test2;")

dbGetQuery(mydb, "DROP TABLE IF EXISTS test2;")

dbGetQuery(mydb,"CREATE TABLE test2 (
  score INTEGER UNIQUE, 
  student TEXT
  );")

dbGetQuery(mydb,"INSERT INTO test2 (score, student) VALUES ('1', 'A');")

#Error because of unique
dbGetQuery(mydb,"INSERT INTO test2 (score, student) VALUES ('1', 'A');")

#NULL is exempt
dbGetQuery(mydb,"INSERT INTO test2 (score, student) VALUES (NULL, 'A');")
dbGetQuery(mydb,"INSERT INTO test2 (score, student) VALUES (NULL, 'A');")

#EXERCISE 5
#Recreate one of your toy data tables with the constraint that for one of the integer variablesthe default value will be zero. Test your table by inserting some empty values. Display your new tables. Then delete your table.  

# 1) Create new table artists_v2 with row_names is set to have default 0
dbGetQuery(db1, "CREATE TABLE artists_v2 (row_names INTEGER DEFAULT 0, artist_name TEXT, role TEXT, year_of_birth INTEGER, country TEXT, zip_code TEXT);")

# 2) Test artists_v2 with some values
dbGetQuery(db1, "INSERT INTO artists_v2 (row_names, artist_name, role, year_of_birth, country, zip_code) VALUES ('1', 'example1', 'rapper', '2000', 'USA', '001');")
dbGetQuery(db1, "INSERT INTO artists_v2 (artist_name, role, year_of_birth, country, zip_code) VALUES ('example2', 'rapper', '2001', 'Korea', '002');")

# 3) Display artists_v2
dbGetQuery(db1, "SELECT * FROM artists_v2")

# 4) Delete artists_v2
dbGetQuery(db1, "DROP TABLE artists_v2")
dbListTables(db1)

```


# Adding a column with a default value
```{r}
#Add a column with default value 1
dbGetQuery(mydb, "ALTER TABLE studentAssessment ADD email INTEGER DEFAULT 1 ")

dbGetQuery(mydb, "SELECT * FROM studentAssessment LIMIT 10;")

#Delete a column
dbGetQuery(mydb, "ALTER TABLE studentAssessment DROP COLUMN email;")

#EXERCISE 6
#Add a column to one of your toy data tables with a default value of 3. Display your new table. Delete this column.

# 1) Add a column "type"
dbGetQuery(db1, "ALTER TABLE albums ADD type INTEGER DEFAULT 3;")

# 2) Display the new column "type"
dbGetQuery(db1, "SELECT * FROM albums LIMIT 10;")

# 3) Delete the new column "type" and display the table 
dbGetQuery(db1, "ALTER TABLE albums DROP COLUMN type;")
dbGetQuery(db1, "SELECT * FROM albums LIMIT 10;")

```


# ID Columns
```{r}
dbGetQuery(mydb,"CREATE TABLE test3 (
  id INTEGER AUTO_INCREMENT PRIMARY KEY, #Not standard syntax
  score INTEGER, 
  student TEXT
  );")

dbGetQuery(mydb,"INSERT INTO test3 (score, student) VALUES (1, 'A');")
dbGetQuery(mydb,"INSERT INTO test3 (score, student) VALUES (5, 'B');")

dbGetQuery(mydb, "SELECT * FROM test3;")

dbGetQuery(mydb, "DROP TABLE IF EXISTS test3;")

#EXERCISE 7
#Create a new table with four variables and a primary key that is a sequential id value.

dbGetQuery(db1, "CREATE TABLE test (id INTEGER AUTO_INCREMENT PRIMARY KEY, type TEXT, score INTEGER, name TEXT, cost INTEGER);")
dbListTables(db1)

```

## Filtering (WHERE)
```{r}
dbGetQuery(mydb, "SELECT id_student, date_submitted FROM studentAssessment WHERE date_submitted > 550 ORDER BY date_submitted DESC;")

#OR Statement
dbGetQuery(mydb, "SELECT id_student, date_submitted FROM studentAssessment WHERE date_submitted > 550 OR date_submitted < 2 ORDER BY date_submitted DESC;")

#AND Statement
dbGetQuery(mydb, "SELECT id_student, date_submitted FROM studentAssessment WHERE date_submitted > 550 AND id_student = 325750 ORDER BY date_submitted DESC;")

#LIKE
dbGetQuery(mydb, "SELECT id_student, gender, region FROM studentInfo WHERE region LIKE '%Region%';")

#Begin with 'Region'
dbGetQuery(mydb, "SELECT id_student, gender, region FROM studentInfo WHERE region LIKE 'Region%';")

#End with 'Region'
dbGetQuery(mydb, "SELECT id_student, gender, region FROM studentInfo WHERE region LIKE '%Region';")

#'c' is the second letter
dbGetQuery(mydb, "SELECT id_student, gender, region FROM studentInfo WHERE region LIKE '_c%';")

#IN
dbGetQuery(mydb, "SELECT id_student, gender, region FROM studentInfo WHERE region IN ('Wales','Ireland');")

#EXERCISE 8
#Query one of your original toy data tables, for two different conditions. 

# 1) Filter with AND statement 
dbGetQuery(db1, "SELECT album_title, num_of_sales, rolling_stone_critic FROM albums WHERE num_of_sales > 999900 AND rolling_stone_critic > 4 ORDER BY num_of_sales;")

# 2) Filter with LIKE statement 
dbGetQuery(db1, "SELECT album_title, genre, num_of_sales FROM albums WHERE genre LIKE '%Heavy Metal%';")

```

## Removing Duplicates
```{r}
dbGetQuery(mydb, "SELECT DISTINCT region FROM studentInfo;")

dbGetQuery(mydb, "SELECT DISTINCT region, gender FROM studentInfo;")

#EXERCISE 9
#Insert a duplicate row into one of your toy data tables. Then query the table without including duplicates.

# 1) Insert a duplicate row into artists_v1 and display it 
dbGetQuery(db1, "INSERT INTO artists_v1 (row_names, artist_name, role, year_of_birth, country, zip_code) VALUES ('1', 'Kameko Nelson', 'female voice', '1981', 'Slovenia', '6914');")
dbGetQuery(db1, "SELECT * FROM artists_v1 ORDER BY row_names LIMIT 5;")

# 2) Query the table without including duplicates
dbGetQuery(db1, "SELECT DISTINCT artist_name, role FROM artists_v1;")

```

## Conditional Expressions (non-standard)
```{r}
dbGetQuery(mydb, "CREATE TABLE booltest (a INTEGER, b INTEGER);")
dbGetQuery(mydb, "INSERT INTO booltest VALUES (1, 0);")
dbGetQuery(mydb, "SELECT * FROM booltest;")

dbGetQuery(mydb,"SELECT
  CASE WHEN a THEN 'true' ELSE 'false' END as boolA,
  CASE WHEN b THEN 'true' ELSE 'false' END as boolB
  FROM booltest")

dbGetQuery(mydb,"SELECT
  CASE a WHEN 1 THEN 'true' ELSE 'false' END as boolA,
  CASE b WHEN 1 THEN 'true' ELSE 'false' END as boolB
  FROM booltest")
```

#Relationships (JOIN) - *Slide*
```{r}

#Create two tables with matches and join them

dbGetQuery(mydb, "CREATE TABLE left_table (id INTEGER, description TEXT);")
dbGetQuery(mydb, "CREATE TABLE right_table (id INTEGER, description TEXT);")

dbGetQuery(mydb, "INSERT INTO left_table VALUES ( 1, 'left 01');")
dbGetQuery(mydb, "INSERT INTO left_table VALUES ( 2, 'left 02');")
dbGetQuery(mydb, "INSERT INTO left_table VALUES ( 3, 'left 03');")
dbGetQuery(mydb, "INSERT INTO left_table VALUES ( 4, 'left 04');")
dbGetQuery(mydb, "INSERT INTO left_table VALUES ( 5, 'left 05');")
dbGetQuery(mydb, "INSERT INTO left_table VALUES ( 6, 'left 06');")
dbGetQuery(mydb, "INSERT INTO left_table VALUES ( 7, 'left 07');")
dbGetQuery(mydb, "INSERT INTO left_table VALUES ( 8, 'left 08');")
dbGetQuery(mydb, "INSERT INTO left_table VALUES ( 9, 'left 09');")

dbGetQuery(mydb, "INSERT INTO right_table VALUES ( 6, 'left 06');")
dbGetQuery(mydb, "INSERT INTO right_table VALUES ( 7, 'left 07');")
dbGetQuery(mydb, "INSERT INTO right_table VALUES ( 8, 'left 08');")
dbGetQuery(mydb, "INSERT INTO right_table VALUES ( 9, 'left 09');")
dbGetQuery(mydb, "INSERT INTO right_table VALUES ( 10, 'left 10');")
dbGetQuery(mydb, "INSERT INTO right_table VALUES ( 11, 'left 11');")
dbGetQuery(mydb, "INSERT INTO right_table VALUES ( 12, 'left 12');")
dbGetQuery(mydb, "INSERT INTO right_table VALUES ( 13, 'left 13');")
dbGetQuery(mydb, "INSERT INTO right_table VALUES ( 14, 'left 14');")

dbGetQuery(mydb, "SELECT * FROM left_table;")
dbGetQuery(mydb, "SELECT * FROM right_table;")

dbGetQuery(mydb,"SELECT l.description AS left_table, r.description AS right_table 
           FROM left_table AS l 
           JOIN right_table AS r ON l.id = r.id")

dbGetQuery(mydb,"SELECT l.description AS left_table, r.description AS right_table 
           FROM left_table AS l 
           RIGHT JOIN right_table AS r ON l.id = r.id")

dbGetQuery(mydb,"SELECT l.description AS left_table, r.description AS right_table 
           FROM left_table AS l 
           LEFT JOIN right_table AS r ON l.id = r.id")

#Union
dbGetQuery(mydb, "SELECT * FROM left_table
           UNION
           SELECT * FROM right_table;")


#EXERCISE 10
# Create a common id variable in your two toy data tables. Then join those tables so that your query returns all the values from one table and only those that match from the other.

# 1) Create artists5 table and albums5 table by selecting 5 values from artists_v1 and albums table (if not creating new tables with each of 5 values, the original 2 tables would have too many values to include) 
dbGetQuery(db1, "CREATE TABLE artists5 (id INTEGER, artist_name TEXT);")
dbGetQuery(db1, "CREATE TABLE albums5 (id INTEGER, album_title TEXT);")

dbGetQuery(db1, "INSERT INTO artists5 (id, artist_name) VALUES (1, 'Kameko Nelson');")
dbGetQuery(db1, "INSERT INTO artists5 (id, artist_name) VALUES (2, 'Sacha Cameron');")
dbGetQuery(db1, "INSERT INTO artists5 (id, artist_name) VALUES (3, 'Thane E. Oliver');")
dbGetQuery(db1, "INSERT INTO artists5 (id, artist_name) VALUES (4, 'Cole E. Joseph');")
dbGetQuery(db1, "INSERT INTO artists5 (id, artist_name) VALUES (5, 'Sonia U. Jensen');")

dbGetQuery(db1, "INSERT INTO albums5 (id, album_title) VALUES (3, 'Go! Technical The Analyst');")
dbGetQuery(db1, "INSERT INTO albums5 (id, album_title) VALUES (4, 'Repeat White Pleasant Matteo');")
dbGetQuery(db1, "INSERT INTO albums5 (id, album_title) VALUES (5, 'Guilty Words');")
dbGetQuery(db1, "INSERT INTO albums5 (id, album_title) VALUES (6, 'House Of Love');")
dbGetQuery(db1, "INSERT INTO albums5 (id, album_title) VALUES (7, 'Yo No Puedo Several Of');")

dbGetQuery(db1, "SELECT * FROM artists5;")
dbGetQuery(db1, "SELECT * FROM albums5;")

# 2) Display the joined table 
dbGetQuery(db1,"SELECT l.artist_name AS artists5, r.album_title AS albums5 
           FROM artists5 AS l 
           LEFT JOIN albums5 AS r ON l.id = r.id")

```
```{r}
#Now disconnect from your database
dbDisconnect(mydb)

#Then retunr to your AWS console and:

#1. Click on "Actions" and then "Stop"
#2. Do NOT make a snapshot
#3 Click on "Actions" again and click "Delete"
#4. Unclick "Make a final snapshot"
#5. Clicl "I acknowledge that upon instance deletion, automated backups, including system snapshots and point-in-time recovery, will no longer be available."
#6. Type "delete me" into the field

#Failure to follow these steps could result in charges to your credit card.


```

