---
title: "sql-workshop"
author: "Jiasheng Yu"
output: html_document
---

Before you follow the directions below, please take a screenshot of your AWS console showing the running database and upload it to your repo.

## Connect to AWS MySQL Database
```{r}
#install.packages("DBI", "RMySQL")
#install.packages('RMySQL')

library(DBI)
library(RMySQL)

db_user <- 'admin'
db_password <- 'testsql!'
db_name <- 'oudb'
db_host <- 'database-1.chp9yqelef9i.us-east-2.rds.amazonaws.com'
db_port <- 3306

mydb <- dbConnect(MySQL(), user = db_user, password = db_password, dbname = db_name, host = db_host, port = db_port)

summary(mydb)
```

## Load OU Data
```{r}
#Student demographic data
studentInfo <- read.csv("studentInfo.csv", header = TRUE)
#Student assessment data
studentAssessment <- read.csv("studentAssessment.csv", header = TRUE)
#Course data
courses <- read.csv("courses.csv", header = TRUE)
studentRegistration <- read.csv("studentRegistration.csv", header = TRUE)
```

## Write data to the DB using the DBI package
```{r}
#List the tables in the DB - should be zero
dbListTables(mydb)

#Write a new table to the DB
dbWriteTable(mydb, "studentInfo", studentInfo)
dbWriteTable(mydb, "studentAssessment", studentAssessment)
dbWriteTable(mydb, "courses", courses)
dbWriteTable(mydb, "studentRegistration", studentRegistration)

#List tables to see that table was added
#dbListTables(mydb)

#Read a particular table
# dbReadTable(mydb, 'studentInfo')

#EXERCISE 1
#Make two toy data sets with at least three variables and at least 30 rows each in them. Have a mix of numeric and character variables. Transfer these dataframes to your SQL database using the DBI commands. Name the tables whatever you like.
library(dplyr)

toy1<-select(studentInfo,code_module,code_presentation,id_student,gender,num_of_prev_attempts,studied_credits,disability,final_result)

toy1<-toy1[c(1:30),]

toy2<-select(studentInfo,id_student,gender,region,highest_education,imd_band,age_band)

toy2<-toy2[c(1:30),]

dbWriteTable(mydb,"stu_grades",toy1)
dbWriteTable(mydb,"stu_info_subset",toy2)

dbListTables(mydb)
```

## Getting into SQL - READING
```{r}
#Query a portion of the database (always returns dataframe)
# dbGetQuery(mydb, "SELECT * FROM studentInfo LIMIT 10;")
# 
# dbGetQuery(mydb, "SELECT * FROM studentInfo ORDER BY id_student LIMIT 10;")
# 
# dbGetQuery(mydb, "SELECT id_student, gender FROM studentInfo ORDER BY id_student DESC LIMIT 10;") #Order listed will be reflected in order in table
# 
# dbGetQuery(mydb, "SELECT id_student AS 'Student ID', gender FROM studentInfo LIMIT 10;") #SQL Standard says quotes for literal strings and double quotes for everything else but that conflicts with R
# 
# #Count the number of rows
# dbGetQuery(mydb, "SELECT COUNT(*) FROM studentAssessment;")
# 
# #Using a WHERE statement on all columns
# dbGetQuery(mydb, "SELECT COUNT(*) FROM studentAssessment WHERE score > 50;")
# 
# #Using a WHERE statement on a single column (will not include missing data)
# dbGetQuery(mydb, "SELECT COUNT(score) FROM studentAssessment WHERE score > 50;")
# 
# #Using an AND statement
# dbGetQuery(mydb, "SELECT COUNT(*) FROM studentAssessment WHERE score > 50 AND id_assessment = '1752';")

#EXERCISE 2
#Read one of your toy data tables, make sure the output is ordered in descending order, you rename one of the variables and the output is limited to the first 20 rows.

dbGetQuery(mydb,"
                  SELECT  code_module,
                          code_presentation,
                          id_student AS stu_id,
                          gender,
                          num_of_prev_attempts,
                          studied_credits,
                          disability,
                          final_result
                  FROM    stu_grades
                  ORDER BY
                          stu_id
                  DESC
                  LIMIT 20;
                ")
#Read the other table according to a condition of one of the variables.

dbGetQuery(mydb,"
                  SELECT  *
                  FROM    stu_info_subset
                  WHERE   gender = 'M';
                ")

```

## Getting into SQL - UPDATING
```{r}
# #Count rows
# dbGetQuery(mydb, "SELECT COUNT(*) FROM studentAssessment;")
# 
# #Add a row
# dbGetQuery(mydb, "INSERT INTO studentAssessment (id_assessment, id_student, date_submitted, is_banked, score) VALUES ('00001', '1', '20', '0', '50');")
# 
# #Count rows again
# dbGetQuery(mydb, "SELECT COUNT(*) FROM studentAssessment;")
# 
# #View inserted row
# dbGetQuery(mydb, "SELECT * FROM studentAssessment ORDER BY id_student LIMIT 10;")
# 
# #Add a row with missing values
# dbGetQuery(mydb, "INSERT INTO studentAssessment (id_assessment, id_student, date_submitted) VALUES ('00001', '1', '20');")
# 
# #View inserted row
# dbGetQuery(mydb, "SELECT * FROM studentAssessment ORDER BY id_student LIMIT 10;")
# 
# #Update a row
# dbGetQuery(mydb, "UPDATE studentAssessment SET score = '20' WHERE id_student = 1;")
# 
# dbGetQuery(mydb, "SELECT id_student, score FROM studentAssessment ORDER BY id_student LIMIT 10;")
# 
# #Update a row with NULL
# dbGetQuery(mydb, "UPDATE studentAssessment SET score = 'NULL' WHERE id_student = 6516;")
# 
# #Delete a row (destructive)
# dbGetQuery(mydb, "DELETE FROM studentAssessment WHERE id_student = 1;")
# 
# dbGetQuery(mydb, "SELECT * FROM studentAssessment ORDER BY id_student LIMIT 10;")

#EXERCISE 3
#Insert a new row in one of your toy data tables leaving one variable empty. Change one value in your other table. Display your new tables. Delete the row you edited and the row you inserted. 

#insert
dbGetQuery(mydb,"
                  INSERT INTO stu_grades (code_module,
                                          code_presentation,
                                          id_student,
                                          gender,
                                          num_of_prev_attempts,
                                          studied_credits,
                                          disability,
                                          final_result) 
                              VALUES     ('BBB',
                                          '2021J',
                                           1,
                                          'M',
                                           10,
                                           100,
                                           NULL ,
                                           'Pass');")
dbGetQuery(mydb," SELECT * FROM stu_grades; ")
#update

dbGetQuery(mydb,"
           
                  UPDATE  stu_info_subset
                  set     gender = 'F'
                  WHERE   id_student = 11391;
           
           ")
dbGetQuery(mydb," SELECT * FROM stu_info_subset WHERE id_student = 11391;")

#delete
dbGetQuery(mydb," DELETE
                  FROM   stu_grades
                  WHERE  disability IS NULL;")
dbGetQuery(mydb," SELECT * FROM stu_grades; ")

dbGetQuery(mydb," DELETE
                  FROM   stu_info_subset
                  WHERE  id_student = 11391;")
dbGetQuery(mydb," SELECT * FROM stu_info_subset WHERE id_student = 11391;")

```

## Add/Deleting Table
```{r}
#Creating a new table in SQL
# dbGetQuery(mydb,"CREATE TABLE test (
#   score INTEGER, 
#   student TEXT
#   );")
# 
# dbListTables(mydb)
# 
# #Inserting data into the table
# dbGetQuery(mydb, "INSERT INTO test VALUES ( 10, 'Amy' );")
# dbGetQuery(mydb, "INSERT INTO test VALUES ( 11, 'Jen' );")
# dbGetQuery(mydb, "INSERT INTO test VALUES ( 9, 'Frank' );")
# 
# dbGetQuery(mydb, "SELECT * FROM test;")
# 
# #Inserting a NULL row
# dbGetQuery(mydb, "INSERT INTO test DEFAULT VALUES;") #Will not work use instead:
# 
# dbGetQuery(mydb,"INSERT INTO test (score, student) SELECT score, id_student FROM studentAssessment;")
# 
# #Delete a table
# dbGetQuery(mydb, "DROP TABLE test;")
# 
# dbGetQuery(mydb, "SELECT * FROM test;") #This should produce an error since your table no longer exists
# 
# #Delete a table if it exists
# dbGetQuery(mydb, "DROP TABLE IF EXISTS test;") #No error since it is only if it exists

#EXERCISE 4
#Create a table that is exactly the same as your first toy data table but this time use SQL commands. Display your new table. Then delete the original table. 

#create

dbGetQuery(mydb," 
                  CREATE TABLE new_stu_grades     
                                (code_module TEXT,
                                 code_presentation TEXT,
                                 id_student INTEGER,
                                 gender TEXT,
                                 num_of_prev_attempts INTEGER,
                                 studied_credits INTEGER,
                                 disability TEXT,
                                 final_result TEXT) 
                
           ")

dbGetQuery(mydb," 
                  INSERT INTO new_stu_grades(
                  
                          code_module,
                          code_presentation,
                          id_student,
                          gender,
                          num_of_prev_attempts,
                          studied_credits,
                          disability,
                          final_result
                          )
                          
                  SELECT  code_module,
                          code_presentation,
                          id_student AS stu_id,
                          gender,
                          num_of_prev_attempts,
                          studied_credits,
                          disability,
                          final_result
                  FROM    stu_grades;")
 
dbGetQuery(mydb,"SELECT * FROM new_stu_grades;")
#delete table
dbGetQuery(mydb,"DROP TABLE stu_grades;")
#dbGetQuery(mydb,"SELECT * FROM  stu_grades;") -- doesn't exist


```

# NULL Value
```{r}
#NULL is a state (similar to R), represents the lack of a value. But is not compatible with R backend so this code doesn't work as part of dbGetQuery()

#This doesn't work because NULL is not a value
# SELECT * FROM test WHERE score = NULL;

#Instead use
# SELECT * FROM test WHERE score is NULL;

```

# Constraints
```{r}
#Create table where student column *cannot* be NULL
# dbGetQuery(mydb,"CREATE TABLE test2 (
#   score INTEGER, 
#   student TEXT NOT NULL
#   );")
# 
# dbGetQuery(mydb, "DROP TABLE IF EXISTS test2;")
# 
# dbGetQuery(mydb,"CREATE TABLE test2 (
#    score INTEGER DEFAULT 0, 
#    student TEXT
#    );")
# 
# dbGetQuery(mydb,"INSERT INTO test2 (score, student) VALUES ('1', 'A');")
# dbGetQuery(mydb,"INSERT INTO test2 (student) VALUES ('B');")
# 
# dbGetQuery(mydb, "SELECT * FROM test2;")
# 
# dbGetQuery(mydb, "DROP TABLE IF EXISTS test2;")
# 
# dbGetQuery(mydb,"CREATE TABLE test2 (
#   score INTEGER UNIQUE, 
#   student TEXT
#   );")
# 
# dbGetQuery(mydb,"INSERT INTO test2 (score, student) VALUES ('1', 'A');")
# 
# #Error because of unique
# dbGetQuery(mydb,"INSERT INTO test2 (score, student) VALUES ('1', 'A');")
# 
# #NULL is exempt
# dbGetQuery(mydb,"INSERT INTO test2 (score, student) VALUES (NULL, 'A');")
# dbGetQuery(mydb,"INSERT INTO test2 (score, student) VALUES (NULL, 'A');")

#EXERCISE 5
#Recreate one of your toy data tables with the constraint that for one of the integer variablesthe default value will be zero. Test your table by inserting some empty values. Display your new tables. Then delete your table. 

dbGetQuery(mydb,"
           
                CREATE TABLE new_stu_info_subset(
                          id_student INTEGER DEFAULT 0,
                          gender TEXT,
                          region TEXT,
                          highest_education TEXT,
                          imd_band TEXT,
                          age_band TEXT
              
                
                );
                            
           ")
dbListTables(mydb)
dbGetQuery(mydb,"
           
                INSERT INTO new_stu_info_subset(
                          
                          gender,
                          region,
                          highest_education,
                          imd_band ,
                          age_band 
                
                )
                VALUES
                ('M','Scotland','HE Qualification','1','2'),
                ('F','Scotland','HE Qualification','3','4')
                
                ")
dbGetQuery(mydb,"
           
                SELECT * FROM new_stu_info_subset;
           
           ")
dbGetQuery(mydb,"DROP TABLE new_stu_info_subset;")
dbListTables(mydb)

```


# Adding a column with a default value
```{r}
# #Add a column with default value 1
# dbGetQuery(mydb, "ALTER TABLE studentAssessment ADD email INTEGER DEFAULT 1 ")
# 
# dbGetQuery(mydb, "SELECT * FROM studentAssessment LIMIT 10;")
# 
# #Delete a column
# dbGetQuery(mydb, "ALTER TABLE studentAssessment DROP COLUMN email;")

#EXERCISE 6
#Add a column to one of your toy data tables with a default value of 3. Display your new table. Delete this column.

dbGetQuery(mydb,"
                  ALTER TABLE stu_info_subset ADD new_col INTEGER DEFAULT 3;
           ")
dbGetQuery(mydb," SELECT * FROM stu_info_subset")
dbGetQuery(mydb," ALTER TABLE stu_info_subset DROP new_col; ")
dbGetQuery(mydb," SELECT * FROM stu_info_subset;")

```


# ID Columns
```{r}
# dbGetQuery(mydb,"CREATE TABLE test3 (
#   id INTEGER AUTO_INCREMENT PRIMARY KEY, #Not standard syntax
#   score INTEGER, 
#   student TEXT
#   );")
# 
# dbGetQuery(mydb,"INSERT INTO test3 (score, student) VALUES (1, 'A');")
# dbGetQuery(mydb,"INSERT INTO test3 (score, student) VALUES (5, 'B');")
# 
# dbGetQuery(mydb, "SELECT * FROM test3;")
# 
# dbGetQuery(mydb, "DROP TABLE IF EXISTS test3;")

#EXERCISE 7
#Create a new table with four variables and a primary key that is a sequential id value.
dbGetQuery(mydb,"
                  CREATE TABLE new_table (
                         id INTEGER AUTO_INCREMENT PRIMARY KEY,
                         name TEXT,
                         email TEXT,
                         address TEXT
                  );
           
           
           ")
dbListTables(mydb)

dbGetQuery(mydb," INSERT INTO new_table (name,email,address) 
                  VALUES('A','A@email.com','Address: A'),
                        ('B','B@email.com','Address: B')")
dbGetQuery(mydb,"SELECT * FROM new_table;")
```

## Filtering (WHERE)
```{r}
# dbGetQuery(mydb, "SELECT id_student, date_submitted FROM studentAssessment WHERE date_submitted > 550 ORDER BY date_submitted DESC;")
# 
# #OR Statement
# dbGetQuery(mydb, "SELECT id_student, date_submitted FROM studentAssessment WHERE date_submitted > 550 OR date_submitted < 2 ORDER BY date_submitted DESC;")
# 
# #AND Statement
# dbGetQuery(mydb, "SELECT id_student, date_submitted FROM studentAssessment WHERE date_submitted > 550 AND id_student = 325750 ORDER BY date_submitted DESC;")
# 
# #LIKE
# dbGetQuery(mydb, "SELECT id_student, gender, region FROM studentInfo WHERE region LIKE '%Region%';")
# 
# #Begin with 'Region'
# dbGetQuery(mydb, "SELECT id_student, gender, region FROM studentInfo WHERE region LIKE 'Region%';")
# 
# #End with 'Region'
# dbGetQuery(mydb, "SELECT id_student, gender, region FROM studentInfo WHERE region LIKE '%Region';")
# 
# #'c' is the second letter
# dbGetQuery(mydb, "SELECT id_student, gender, region FROM studentInfo WHERE region LIKE '_c%';")
# 
# #IN
# dbGetQuery(mydb, "SELECT id_student, gender, region FROM studentInfo WHERE region IN ('Wales','Ireland');")

#EXERCISE 8
#Query one of your original toy data tables, for two different conditions. 

dbGetQuery(mydb,"
                  SELECT  *
                  FROM    stu_info_subset
                  WHERE   highest_education LIKE '_e%';
           
           
           ")

dbGetQuery(mydb,"
                  SELECT  *
                  FROM    stu_info_subset
                  WHERE   region LIKE '%Region%'
                  	AND   imd_band IN ('30-40%', '70-80%');
           
           ")

```

## Removing Duplicates
```{r}
# dbGetQuery(mydb, "SELECT DISTINCT region FROM studentInfo;")
# 
# dbGetQuery(mydb, "SELECT DISTINCT region, gender FROM studentInfo;")

#EXERCISE 9
#Insert a duplicate row into one of your toy data tables. Then query the table without including duplicates.

dbGetQuery(mydb,"SELECT * FROM stu_info_subset;")
dbGetQuery(mydb,"
                  INSERT INTO stu_info_subset
                  VALUES  (2,28400,'F','Scotland','HE Qualification','20-30%','35-55')
           
           ")

dbGetQuery(mydb," SELECT DISTINCT * FROM stu_info_subset;")

```

## Conditional Expressions (non-standard)
```{r}
# dbGetQuery(mydb, "CREATE TABLE booltest (a INTEGER, b INTEGER);")
# dbGetQuery(mydb, "INSERT INTO booltest VALUES (1, 0);")
# dbGetQuery(mydb, "SELECT * FROM booltest;")
# 
# dbGetQuery(mydb,"SELECT
#   CASE WHEN a THEN 'true' ELSE 'false' END as boolA,
#   CASE WHEN b THEN 'true' ELSE 'false' END as boolB
#   FROM booltest")
# 
# dbGetQuery(mydb,"SELECT
#   CASE a WHEN 1 THEN 'true' ELSE 'false' END as boolA,
#   CASE b WHEN 1 THEN 'true' ELSE 'false' END as boolB
#   FROM booltest")
```

#Relationships (JOIN) - *Slide*
```{r}

#Create two tables with matches and join them

# dbGetQuery(mydb, "CREATE TABLE left_table (id INTEGER, description TEXT);")
# dbGetQuery(mydb, "CREATE TABLE right_table (id INTEGER, description TEXT);")
# 
# dbGetQuery(mydb, "INSERT INTO left_table VALUES ( 1, 'left 01');")
# dbGetQuery(mydb, "INSERT INTO left_table VALUES ( 2, 'left 02');")
# dbGetQuery(mydb, "INSERT INTO left_table VALUES ( 3, 'left 03');")
# dbGetQuery(mydb, "INSERT INTO left_table VALUES ( 4, 'left 04');")
# dbGetQuery(mydb, "INSERT INTO left_table VALUES ( 5, 'left 05');")
# dbGetQuery(mydb, "INSERT INTO left_table VALUES ( 6, 'left 06');")
# dbGetQuery(mydb, "INSERT INTO left_table VALUES ( 7, 'left 07');")
# dbGetQuery(mydb, "INSERT INTO left_table VALUES ( 8, 'left 08');")
# dbGetQuery(mydb, "INSERT INTO left_table VALUES ( 9, 'left 09');")
# 
# dbGetQuery(mydb, "INSERT INTO right_table VALUES ( 6, 'left 06');")
# dbGetQuery(mydb, "INSERT INTO right_table VALUES ( 7, 'left 07');")
# dbGetQuery(mydb, "INSERT INTO right_table VALUES ( 8, 'left 08');")
# dbGetQuery(mydb, "INSERT INTO right_table VALUES ( 9, 'left 09');")
# dbGetQuery(mydb, "INSERT INTO right_table VALUES ( 10, 'left 10');")
# dbGetQuery(mydb, "INSERT INTO right_table VALUES ( 11, 'left 11');")
# dbGetQuery(mydb, "INSERT INTO right_table VALUES ( 12, 'left 12');")
# dbGetQuery(mydb, "INSERT INTO right_table VALUES ( 13, 'left 13');")
# dbGetQuery(mydb, "INSERT INTO right_table VALUES ( 14, 'left 14');")
# 
# dbGetQuery(mydb, "SELECT * FROM left_table;")
# dbGetQuery(mydb, "SELECT * FROM right_table;")
# 
# dbGetQuery(mydb,"SELECT l.description AS left_table, r.description AS right_table 
#            FROM left_table AS l 
#            JOIN right_table AS r ON l.id = r.id")
# 
# dbGetQuery(mydb,"SELECT l.description AS left_table, r.description AS right_table 
#            FROM left_table AS l 
#            RIGHT JOIN right_table AS r ON l.id = r.id")
# 
# dbGetQuery(mydb,"SELECT l.description AS left_table, r.description AS right_table 
#            FROM left_table AS l 
#            LEFT JOIN right_table AS r ON l.id = r.id")
# 
# #Union
# dbGetQuery(mydb, "SELECT * FROM left_table
#            UNION
#            SELECT * FROM right_table;")


#EXERCISE 10
# Create a common id variable in your two toy data tables. Then join those tables so that your query returns all the values from one table and only those that match from the other. 

#I create the common id from the beginning,so I just use them
dbListTables(mydb)
dbGetQuery(mydb,"DESC new_stu_grades;")
dbGetQuery(mydb,"SELECT * FROM new_stu_grades;")
dbGetQuery(mydb,"DESC stu_info_subset")
dbGetQuery(mydb,"SELECT * FROM stu_info_subset;")

# dbGetQuery(mydb," select distinct a.id_student,b.id_student from new_stu_grades a
#                   left join
#                   stu_info_subset b
#                   on a.id_student = b.id_student
#                   order by a.id_student desc
#            ")
dbGetQuery(mydb," 
                  SELECT  a.*,
                          b.*
                  FROM    new_stu_grades a
                  LEFT JOIN
                          stu_info_subset b
                      ON    
                          a.id_student = b.id_student
           
           
           
           ")
```

```{r}
#Now disconnect from your database
dbDisconnect(mydb)

#Then retunr to your AWS console and:

#1. Click on "Actions" and then "Stop"
#2. Do NOT make a snapshot
#3 Click on "Actions" again and click "Delete"
#4. Unclick "Make a final snapshot"
#5. Clicl "I acknowledge that upon instance deletion, automated backups, including system snapshots and point-in-time recovery, will no longer be available."
#6. Type "delete me" into the field

#Failure to follow these steps could result in charges to your credit card.


```

