---
title: "sql-workshop"
author: "Guotai Sun"
output: html_document
---

Before you follow the directions below, please take a screenshot of your AWS console showing the running database and upload it to your repo.

## Connect to AWS MySQL Database
```{r}
install.packages("DBI", "RMySQL")

library(DBI)
library(RMySQL)

db_user <- 'gs3144'
db_password <- 'Michael840193'
db_name <- 'oudb'
db_host <- 'database-1.cl7ipplfercw.us-east-2.rds.amazonaws.com'
db_port <- 3306

mydb <- dbConnect(MySQL(), user = db_user, password = db_password, dbname = db_name, host = db_host, port = db_port)

summary(mydb)
```

## Load OU Data
```{r}
PI <- read.csv("PI.csv", header = TRUE)

PA <- read.csv("PA.csv", header = TRUE)

```

## Write data to the DB using the DBI package
```{r}
dbWriteTable(mydb, "PI", PI)

dbWriteTable(mydb, "PA", PA)

#Read a particular table
dbReadTable(mydb, 'PI')

dbReadTable(mydb, 'PA')

#EXERCISE 1
#Make two toy data sets with at least three variables and at least 30 rows each in them. Have a mix of numeric and character variables. Transfer these dataframes to your SQL database using the DBI commands. Name the tables whatever you like.

```

## Getting into SQL - READING
```{r}
#Query a portion of the database (always returns dataframe)
dbGetQuery(mydb, "SELECT * FROM PI LIMIT 20;")

dbGetQuery(mydb, "SELECT * FROM PI ORDER BY id_professor LIMIT 20;")

dbGetQuery(mydb, "SELECT id_professor, gender FROM PI ORDER BY id_professor DESC LIMIT 10;") 

dbGetQuery(mydb, "SELECT id_professor AS 'professor ID', gender FROM PI LIMIT 20;") 

dbGetQuery(mydb, "SELECT COUNT(*) FROM PA;")

dbGetQuery(mydb, "SELECT COUNT(*) FROM PA WHERE date_submitted > 20;")

dbGetQuery(mydb, "SELECT COUNT(date_submitted) FROM PA WHERE date_submitted > 20;")

dbGetQuery(mydb, "SELECT COUNT(*) FROM PA WHERE date_submitted > 20 AND is_banked = '0';")

#EXERCISE 2
#Read one of your toy data tables, make sure the output is ordered in descending order, you rename one of the variables and the output is limited to the first 20 rows.

#Read the other table according to a condition of one of the variables.

```

## Getting into SQL - UPDATING
```{r}
#Count rows
dbGetQuery(mydb, "SELECT COUNT(*) FROM PA;")

#Add a row
dbGetQuery(mydb, "INSERT INTO PA (id_professor,date_submitted,is_banked) VALUES ('00001', '1', '0');")

#Count rows again
dbGetQuery(mydb, "SELECT COUNT(*) FROM PA;")

#View inserted row
dbGetQuery(mydb, "SELECT * FROM PA ORDER BY id_professor LIMIT 20;")

#Add a row with missing values
dbGetQuery(mydb, "INSERT INTO PA (id_professor,date_submitted) VALUES ('00001', '1');")

#View inserted row
dbGetQuery(mydb, "SELECT * FROM PA ORDER BY id_professor LIMIT 20;")

#Update a row
dbGetQuery(mydb, "UPDATE PA SET is_banked = '1' WHERE id_professor = 00001;")

dbGetQuery(mydb, "SELECT id_professor,is_banked FROM PA ORDER BY id_professor LIMIT 20;")

#Update a row with NULL
dbGetQuery(mydb, "UPDATE PA SET is_banked = 'NULL' WHERE id_professor = 65481;")

#Delete a row (destructive)
dbGetQuery(mydb, "DELETE FROM PA WHERE id_professor = 00001;")

dbGetQuery(mydb, "SELECT * FROM PAt ORDER BY id_professor LIMIT 20;")

#EXERCISE 3
#Insert a new row in one of your toy data tables leaving one variable empty. Change one value in your other table. Display your new tables. Delete the row you edited and the row you inserted. 


```

## Add/Deleting Table
```{r}
#Creating a new table in SQL
dbGetQuery(mydb,"CREATE TABLE test (
  id_professor INTEGER,
  date_submitted INTEGER,
  is_banked INTEGER
  );")

#Inserting data into the table
dbGetQuery(mydb, "INSERT INTO test VALUES ( 9998, 19,0 );")
dbGetQuery(mydb, "INSERT INTO test VALUES ( 9999, 20,0 );")
dbGetQuery(mydb, "INSERT INTO test VALUES ( 1000, 25,0 );")

dbGetQuery(mydb, "SELECT * FROM test;")

#Inserting a NULL row

dbGetQuery(mydb,"INSERT INTO test (id_professor,date_submitted,is_banked) SELECT id_professor,date_submitted,is_banked FROM PA;")

#Delete a table
dbGetQuery(mydb, "DROP TABLE test;")

dbGetQuery(mydb, "SELECT * FROM test;") #This should produce an error since your table no longer exists

#Delete a table if it exists
dbGetQuery(mydb, "DROP TABLE IF EXISTS test;") #No error since it is only if it exists

#EXERCISE 4
#Create a table that is exactly the same as your first toy data table but this time use SQL commands. Display your new table. Then delete the original table. 

```
# Constraints
```{r}

dbGetQuery(mydb,"CREATE TABLE test9 (
  date_submitted INTEGER, 
  professor TEXT NOT NULL
  );")

dbGetQuery(mydb, "DROP TABLE IF EXISTS test9;")

dbGetQuery(mydb,"CREATE TABLE test9 (
   date_submitted INTEGER DEFAULT 0, 
   professor TEXT
   );")

dbGetQuery(mydb,"INSERT INTO test9 (date_submitted, professor) VALUES ('19', 'A');")
dbGetQuery(mydb,"INSERT INTO test9 (professor) VALUES ('B');")

dbGetQuery(mydb, "SELECT * FROM test9;")

dbGetQuery(mydb, "DROP TABLE IF EXISTS test9;")

#NULL is exempt
dbGetQuery(mydb,"INSERT INTO test9 (date_submitted, professor) VALUES (NULL, 'A');")
dbGetQuery(mydb,"INSERT INTO test9 (date_submitted, professsor) VALUES (NULL, 'A');")

#EXERCISE 5
#Recreate one of your toy data tables with the constraint that for one of the integer variablesthe default value will be zero. Test your table by inserting some empty values. Display your new tables. Then delete your table.  

```
# Adding a column with a default value
```{r}
#Add a column with default value 1
dbGetQuery(mydb, "ALTER TABLE PA ADD email INTEGER DEFAULT 3 ")

dbGetQuery(mydb, "SELECT * FROM PA LIMIT 20;")

#Delete a column
dbGetQuery(mydb, "ALTER TABLE PA DROP COLUMN email;")

#EXERCISE 6
#Add a column to one of your toy data tables with a default value of 3. Display your new table. Delete this column.
```
# ID Columns
```{r}
dbGetQuery(mydb,"CREATE TABLE test3 (
  id INTEGER AUTO_INCREMENT PRIMARY KEY, 
  age INTEGER,
  score INTEGER,
  is_banked INTEGER,
  professor TEXT
  );")

dbGetQuery(mydb,"INSERT INTO test3 (age,score, is_banked, professor) VALUES (25,90,0, 'A');")
dbGetQuery(mydb,"INSERT INTO test3 (age,score, is_banked, professor) VALUES (30,95,0, 'B');")

dbGetQuery(mydb, "SELECT * FROM test3;")

dbGetQuery(mydb, "DROP TABLE IF EXISTS test3;")

#EXERCISE 7
#Create a new table with four variables and a primary key that is a sequential id value.

```

## Filtering (WHERE)
```{r}
dbGetQuery(mydb, "SELECT id_professor, date_submitted FROM PA WHERE date_submitted < 49 ORDER BY date_submitted DESC;")

#LIKE
dbGetQuery(mydb, "SELECT id_professor, gender, region FROM PI WHERE region LIKE 'South Region';")

#'c' is the second letter
dbGetQuery(mydb, "SELECT id_professor, gender, region FROM PI WHERE region LIKE '_r%';")

#EXERCISE 8
#Query one of your original toy data tables, for two different conditions. 

```

## Removing Duplicates
```{r}
dbGetQuery(mydb, "SELECT DISTINCT region FROM PI;")

dbGetQuery(mydb, "SELECT DISTINCT region, gender FROM PI;")

#EXERCISE 9
#Insert a duplicate row into one of your toy data tables. Then query the table without including duplicates.

```

## Conditional Expressions (non-standard)
```{r}
dbGetQuery(mydb, "CREATE TABLE booltest (a INTEGER, b INTEGER);")
dbGetQuery(mydb, "INSERT INTO booltest VALUES (1, 0);")
dbGetQuery(mydb, "SELECT * FROM booltest;")

dbGetQuery(mydb,"SELECT
  CASE WHEN a THEN 'true' ELSE 'false' END as boolA,
  CASE WHEN b THEN 'true' ELSE 'false' END as boolB
  FROM booltest")

dbGetQuery(mydb,"SELECT
  CASE a WHEN 1 THEN 'true' ELSE 'false' END as boolA,
  CASE b WHEN 1 THEN 'true' ELSE 'false' END as boolB
  FROM booltest")
```

#Relationships (JOIN) - *Slide*
```{r}

dbGetQuery(mydb, "INSERT INTO PA (id_professor,date_submitted,is_banked) VALUES ( 1, 19,0);")
dbGetQuery(mydb, "INSERT INTO PA (id_professor,date_submitted,is_banked) VALUES ( 2, 22,0);")
dbGetQuery(mydb, "INSERT INTO PA (id_professor,date_submitted,is_banked) VALUES ( 3, 25,0);")
dbGetQuery(mydb, "INSERT INTO PA (id_professor,date_submitted,is_banked) VALUES ( 4, 23,0);")
dbGetQuery(mydb, "INSERT INTO PA (id_professor,date_submitted,is_banked) VALUES ( 5, 24,0);")
dbGetQuery(mydb, "INSERT INTO PA (id_professor,date_submitted,is_banked) VALUES ( 6, 18,0);")
dbGetQuery(mydb, "INSERT INTO PA (id_professor,date_submitted,is_banked) VALUES ( 7, 19,0);")
dbGetQuery(mydb, "INSERT INTO PA (id_professor,date_submitted,is_banked) VALUES ( 8, 26,0);")
dbGetQuery(mydb, "INSERT INTO PA (id_professor,date_submitted,is_banked) VALUES ( 9, 24,0);")

dbGetQuery(mydb, "INSERT INTO PI (id_professor,gender,region) VALUES ( 6, 'M', 'NA');")
dbGetQuery(mydb, "INSERT INTO PI (id_professor,gender,region) VALUES ( 7, 'M', 'NA');")
dbGetQuery(mydb, "INSERT INTO PI (id_professor,gender,region) VALUES ( 8, 'M', 'NA');")
dbGetQuery(mydb, "INSERT INTO PI (id_professor,gender,region) VALUES ( 9, 'F', 'NA');")
dbGetQuery(mydb, "INSERT INTO PI (id_professor,gender,region) VALUES ( 10, 'F', 'CN');")
dbGetQuery(mydb, "INSERT INTO PI (id_professor,gender,region) VALUES ( 11, 'F', 'CN');")
dbGetQuery(mydb, "INSERT INTO PI (id_professor,gender,region) VALUES ( 12, 'M', 'CN');")
dbGetQuery(mydb, "INSERT INTO PI (id_professor,gender,region) VALUES ( 13, 'F', 'NA');")
dbGetQuery(mydb, "INSERT INTO PI (id_professor,gender,region) VALUES ( 14, 'M', 'CN');")

dbGetQuery(mydb, "SELECT * FROM PA;")
dbGetQuery(mydb, "SELECT * FROM PI;")

dbGetQuery(mydb,"SELECT PA.id_professor AS PA, PI.id_professor AS PI 
           FROM PA AS PA 
           JOIN PI AS PI ON PA.id_professor = PI.id_professor")

dbGetQuery(mydb,"SELECT PA.id_professor AS PA, PI.id_professor AS PI 
           FROM PA AS PA 
           RIGHT JOIN PI AS PI ON PA.id_professor = PI.id_professor")

dbGetQuery(mydb,"SELECT PA.id_professor AS PA, PI.id_professor AS PI 
           FROM PA AS PA 
           LEFT JOIN PI AS PI ON PA.id_professor = PI.id_professor")

#Union
dbGetQuery(mydb, "SELECT * FROM PA
           UNION
           SELECT * FROM PI;")


#EXERCISE 10
# Create a common id variable in your two toy data tables. Then join those tables so that your query returns all the values from one table and only those that match from the other. 

```
```{r}
#Now disconnect from your database
dbDisconnect(mydb)

#Then retunr to your AWS console and:

#1. Click on "Actions" and then "Stop"
#2. Do NOT make a snapshot
#3 Click on "Actions" again and click "Delete"
#4. Unclick "Make a final snapshot"
#5. Clicl "I acknowledge that upon instance deletion, automated backups, including system snapshots and point-in-time recovery, will no longer be available."
#6. Type "delete me" into the field

#Failure to follow these steps could result in charges to your credit card.


```

