---
title: "sql-workshop"
author: "Charles Lang"
output: html_document
---

Before you follow the directions below, please take a screenshot of your AWS console showing the running database and upload it to your repo.

## Connect to AWS MySQL Database
```{r}
#install.packages("DBI", "RMySQL")

library(DBI)
library(RMySQL)

db_user <- 'admin'
db_password <- 'testsql!'
db_name <- 'outb'
db_host <- 'database-1.crwvhfcv0oya.us-west-2.rds.amazonaws.com'
db_port <- 3306

mydb <- dbConnect(MySQL(), user = db_user, password = db_password, dbname = db_name, host = db_host, port = db_port)

summary(mydb)
```

## Load OU Data
```{r}
#Student demographic data
studentInfo <- read.csv("studentInfo.csv", header = TRUE)
#Student assessment data
studentAssessment <- read.csv("studentAssessment.csv", header = TRUE)
#Course data
courses <- read.csv("courses.csv", header = TRUE)
studentRegistration <- read.csv("studentRegistration.csv", header = TRUE)
```

## Write data to the DB using the DBI package
```{r}
#List the tables in the DB - should be zero
dbListTables(mydb)

#Write a new table to the DB
dbWriteTable(mydb, "studentInfo", studentInfo)
dbWriteTable(mydb, "studentAssessment", studentAssessment)
dbWriteTable(mydb, "courses", courses)
dbWriteTable(mydb, "studentRegistration", studentRegistration)

#List tables to see that table was added
dbListTables(mydb)

#Read a particular table
dbReadTable(mydb, 'studentInfo')

#EXERCISE 1
#Make two toy data sets with at least three variables and at least 30 rows each in them. Have a mix of numeric and character variables. Transfer these dataframes to your SQL database using the DBI commands. Name the tables whatever you like.

#toy1
n1 <- sample(x=1:100,size = 30)
n2 <- sample(x=1:100,size = 30)
n3 <-sample(LETTERS,size=30,replace=TRUE)
toy1 <- data.frame(n1,n2,n3)

#toy2
n1 <- sample(x=1:100,size = 30)
n2 <- sample(x=1:100,size = 30)
n3 <-sample(LETTERS,size=30,replace=TRUE)
toy2 <- data.frame(n1,n2,n3)

#write into the database in mydb
dbWriteTable(mydb, "toy1", toy1)
dbWriteTable(mydb, "toy2", toy2)

#check
dbReadTable(mydb, 'toy1')
dbReadTable(mydb, 'toy2')

```

## Getting into SQL - READING
```{r}
#Query a portion of the database (always returns dataframe)

#write SQL in R, always need dbGetQuery in front of SQL code
#select all, limit to 10 rows
dbGetQuery(mydb, "SELECT * FROM studentInfo LIMIT 10;")

#select all, order by id_student
dbGetQuery(mydb, "SELECT * FROM studentInfo ORDER BY id_student LIMIT 10;")

dbGetQuery(mydb, "SELECT id_student, gender FROM studentInfo ORDER BY id_student DESC LIMIT 10;") #Order listed will be reflected in order in table

#as as rename
dbGetQuery(mydb, "SELECT id_student AS 'Student ID', gender FROM studentInfo LIMIT 10;") #SQL Standard says quotes for literal strings and double quotes for everything else but that conflicts with R

#Count the number of rows
dbGetQuery(mydb, "SELECT COUNT(*) FROM studentAssessment;")

#Using a WHERE statement on all columns
dbGetQuery(mydb, "SELECT COUNT(*) FROM studentAssessment WHERE score > 50;")

#Using a WHERE statement on a single column (will not include missing data)
dbGetQuery(mydb, "SELECT COUNT(score) FROM studentAssessment WHERE score > 50;")

#Using an AND statement
dbGetQuery(mydb, "SELECT COUNT(*) FROM studentAssessment WHERE score > 50 AND id_assessment = '1752';")

#EXERCISE 2
#Read one of your toy data tables, make sure the output is ordered in descending order, you rename one of the variables and the output is limited to the first 20 rows.

#Read the other table according to a condition of one of the variables.

#ANSWER

#toy1 table
#read one table
#ordered in descending order
#rename one of the variables (use as)
#limit to 20 rows (limit 20)
dbGetQuery(mydb, "SELECT n1 as number1, n2, n3 FROM toy1 ORDER BY n1 DESC LIMIT 20;")

#toy2
dbGetQuery(mydb, "SELECT n1 as number11, n2, n3 FROM toy2 ORDER BY n1 DESC LIMIT 20;")


```

## Getting into SQL - UPDATING
```{r}
#Count rows
dbGetQuery(mydb, "SELECT COUNT(*) FROM studentAssessment;")

#Add a row
dbGetQuery(mydb, "INSERT INTO studentAssessment (id_assessment, id_student, date_submitted, is_banked, score) VALUES ('00001', '1', '20', '0', '50');")

#Count rows again
dbGetQuery(mydb, "SELECT COUNT(*) FROM studentAssessment;")

#View inserted row
dbGetQuery(mydb, "SELECT * FROM studentAssessment ORDER BY id_student LIMIT 10;")

#Add a row with missing values
#student_ID =1
dbGetQuery(mydb, "INSERT INTO studentAssessment (id_assessment, id_student, date_submitted) VALUES ('00001', '1', '20');")

#View inserted row
dbGetQuery(mydb, "SELECT * FROM studentAssessment ORDER BY id_student LIMIT 10;")

#Update a row
dbGetQuery(mydb, "UPDATE studentAssessment SET score = '20' WHERE id_student = 1;")

#view the table
dbGetQuery(mydb, "SELECT id_student, score FROM studentAssessment ORDER BY id_student LIMIT 10;")

#Update a row with NULL
dbGetQuery(mydb, "UPDATE studentAssessment SET score = 'NULL' WHERE id_student = 6516;")

#Delete a row (destructive)
#remove student ID=1
dbGetQuery(mydb, "DELETE FROM studentAssessment WHERE id_student = 1;")

#view the table
dbGetQuery(mydb, "SELECT * FROM studentAssessment ORDER BY id_student LIMIT 10;")


#EXERCISE 3
#Insert a new row in one of your toy data tables leaving one variable empty. Change one value in your other table. Display your new tables. Delete the row you edited and the row you inserted. 

#ANSWER

#check all rows
dbGetQuery(mydb, "SELECT COUNT(*) FROM toy1;")

#insert a row into toy1, one variable empty, n3=null
dbGetQuery(mydb, "INSERT INTO toy1 (n1, n2) VALUES ('100', '200');")

#change one value in toy2
dbGetQuery(mydb, "UPDATE toy2 SET n1 = '558' WHERE n3 = 'T';")


#show new tables
dbGetQuery(mydb, "SELECT * FROM toy1 ORDER BY n1 DESC LIMIT 10;")
dbGetQuery(mydb, "SELECT * FROM toy2 ORDER BY n1 DESC LIMIT 10;")


#delete the inserted row in toy1
dbGetQuery(mydb, "DELETE FROM toy1 WHERE n1 = 100 AND n2 = 200;")

#delete the edited row in toy2
dbGetQuery(mydb, "DELETE FROM toy2 WHERE n1 = 558 AND n3 = 'T';")


#display tables
dbGetQuery(mydb, "SELECT * FROM toy1 ORDER BY n1 DESC LIMIT 10;")
dbGetQuery(mydb, "SELECT * FROM toy2 ORDER BY n1 DESC LIMIT 10;")

#check: count toy1 delete insert
dbGetQuery(mydb, "SELECT COUNT(*) FROM toy1;")
```

## Add/Deleting Table
```{r}
#Creating a new table in SQL
dbGetQuery(mydb,"CREATE TABLE test (
  score INTEGER, 
  student TEXT
  );")

dbListTables(mydb)

#Inserting data into the table
dbGetQuery(mydb, "INSERT INTO test VALUES ( 10, 'Amy' );")
dbGetQuery(mydb, "INSERT INTO test VALUES ( 11, 'Jen' );")
dbGetQuery(mydb, "INSERT INTO test VALUES ( 9, 'Frank' );")

dbGetQuery(mydb, "SELECT * FROM test;")

#Inserting a NULL row
dbGetQuery(mydb, "INSERT INTO test DEFAULT VALUES;") 
#^^ Will not work 
#use instead:
dbGetQuery(mydb,"INSERT INTO test (score, student) SELECT score, id_student FROM studentAssessment;")

#Delete a table
dbGetQuery(mydb, "DROP TABLE test;")

dbGetQuery(mydb, "SELECT * FROM test;") 
#This should produce an error since your table no longer exists

#Delete a table if it exists
dbGetQuery(mydb, "DROP TABLE IF EXISTS test;") #No error since it is only if it exists

#EXERCISE 4
#Create a table that is exactly the same as your first toy data table but this time use SQL commands. Display your new table. Then delete the original table. 

#ANWSWER
#create a table in SQL
#dbGetQuery(mydb,"CREATE TABLE toy11(
#           n1 INTEGER,
#           n2 INTEGER,
#           n3 TEXT
#);")

#insert values from table toy1
#dbGetQuery(mydb,"INSERT INTO toy11 (n1, n2, n3) SELECT n1, n2, n3 FROM toy1;")


#insert data into toy11, copy table toy1
dbGetQuery(mydb, "CREATE TABLE toynew1 AS
           SELECT n1,n2,n3
           FROM toy1;")

#check the list in mydb
dbListTables(mydb)

#display the table
dbGetQuery(mydb, "SELECT * FROM toynew1;")

#delete the original table
dbGetQuery(mydb,"DROP TABLE toy1;")

#check if the original table is deleted
dbGetQuery(mydb,"DROP TABLE IF EXISTS toy1;")

```

# NULL Value
```{r}
#NULL is a state (similar to R), represents the lack of a value. But is not compatible with R backend so this code doesn't work as part of dbGetQuery()

#This doesn't work because NULL is not a value
SELECT * FROM test WHERE score = NULL;

#Instead use
SELECT * FROM test WHERE score is NULL;

```

# Constraints
```{r}
#Create table where student column *cannot* be NULL
dbGetQuery(mydb,"CREATE TABLE test2 (
  score INTEGER, 
  student TEXT NOT NULL
  );")

dbGetQuery(mydb, "DROP TABLE IF EXISTS test2;")

#default zero for integer in score
dbGetQuery(mydb,"CREATE TABLE test2 (
   score INTEGER DEFAULT 0, 
   student TEXT
   );")

#insert a row
dbGetQuery(mydb,"INSERT INTO test2 (score, student) VALUES ('1', 'A');")

#give 0 in score as the deadult is set to be zero,(0,8)
dbGetQuery(mydb,"INSERT INTO test2 (student) VALUES ('B');")

#show
dbGetQuery(mydb, "SELECT * FROM test2;")

#drop
dbGetQuery(mydb, "DROP TABLE IF EXISTS test2;")

#unique in the table
dbGetQuery(mydb,"CREATE TABLE test2 (
  score INTEGER UNIQUE, 
  student TEXT
  );")

dbGetQuery(mydb,"INSERT INTO test2 (score, student) VALUES ('1', 'A');")

#Error because of unique
dbGetQuery(mydb,"INSERT INTO test2 (score, student) VALUES ('1', 'A');")

#however, in unique table, NULL will not cause error
#NULL is exempt
dbGetQuery(mydb,"INSERT INTO test2 (score, student) VALUES (NULL, 'A');")
dbGetQuery(mydb,"INSERT INTO test2 (score, student) VALUES (NULL, 'A');")

#EXERCISE 5
#Recreate one of your toy data tables with the constraint that for one of the integer variablesthe default value will be zero. Test your table by inserting some empty values. Display your new tables. Then delete your table.  

#ANSWER
#for integer var, default vale = 0
dbGetQuery(mydb, "CREATE TABLE toynew2(
  n1 INTEGER DEFAULT 0,
  n2 INTEGER,
  n3 TEXT
);")

dbGetQuery(mydb,"INSERT INTO toynew2 (n2, n3) SELECT n2, n3 FROM toy2;")

#copy the table and alter the columb to defult does not work, n1 is not 0
#dbGetQuery(mydb, "CREATE TABLE toynew2 AS
#           SELECT n1,n2,n3
#           FROM toy2;")
#dbGetQuery(mydb, "ALTER TABLE toynew2 ALTER COLUMN n1 SET DEFAULT 0;")

#check table display
dbGetQuery(mydb, "SELECT * FROM toynew2;")

#insert two rows of empty values
dbGetQuery(mydb,"INSERT INTO toynew2 () VALUES ();")
dbGetQuery(mydb,"INSERT INTO toynew2 () VALUES ();")

#display table
dbGetQuery(mydb, "SELECT * FROM toynew2;")

#delete table
dbGetQuery(mydb, "DROP TABLE IF EXISTS toynew2;")

```


# Adding a column with a default value
```{r}
#Add a column with default value 1
dbGetQuery(mydb, "ALTER TABLE studentAssessment ADD email INTEGER DEFAULT 1") 
#notice no ";" here, bc column doesn't exist yet

#display the table
dbGetQuery(mydb, "SELECT * FROM studentAssessment LIMIT 10;")

#Delete a column
dbGetQuery(mydb, "ALTER TABLE studentAssessment DROP COLUMN email;")

#EXERCISE 6
#Add a column to one of your toy data tables with a default value of 3. Display your new table. Delete this column.

#ANSWER
#add a column with default of 3
dbGetQuery(mydb, "ALTER TABLE toy2 ADD ex INTEGER DEFAULT 3")

#display
dbGetQuery(mydb, "SELECT * FROM toy2 LIMIT 10;")

#delete
dbGetQuery(mydb, "ALTER TABLE toy2 DROP COLUMN ex;")

```


# ID Columns
```{r}
#++id
dbGetQuery(mydb,"CREATE TABLE test3 (
  id INTEGER AUTO_INCREMENT PRIMARY KEY, #Not standard syntax
  score INTEGER, 
  student TEXT
  );")

dbGetQuery(mydb,"INSERT INTO test3 (score, student) VALUES (1, 'A');")
dbGetQuery(mydb,"INSERT INTO test3 (score, student) VALUES (5, 'B');")

dbGetQuery(mydb, "SELECT * FROM test3;")

dbGetQuery(mydb, "DROP TABLE IF EXISTS test3;")

#EXERCISE 7
#Create a new table with four variables and a primary key that is a sequential id value.

#aNswer
#a new table with four variables + id = primary key

dbGetQuery(mydb,"CREATE TABLE EX7 (
  id INTEGER AUTO_INCREMENT PRIMARY KEY,
  column1 INTEGER,
  column2 INTEGER,
  column3 TEXT,
  column4 TEXT
  );")

#insert values
dbGetQuery(mydb, "INSERT INTO EX7 (column1, column2, column3, column4) VALUES (1,2,'a','b');")
dbGetQuery(mydb, "INSERT INTO EX7 (column1, column2, column3, column4) VALUES (5,6,'aa','bb');")
dbGetQuery(mydb, "INSERT INTO EX7 (column1, column2, column3, column4) VALUES (3,4,'aaa','bbb');")

#display
dbGetQuery(mydb, "SELECT * FROM EX7;")

#drop table
#dbGetQuery(mydb, "DROP TABLE IF EXISTS EX7;")

```

## Filtering (WHERE)
```{r}
dbGetQuery(mydb, "SELECT id_student, date_submitted FROM studentAssessment WHERE date_submitted > 550 ORDER BY date_submitted DESC;")

#OR Statement
dbGetQuery(mydb, "SELECT id_student, date_submitted FROM studentAssessment WHERE date_submitted > 550 OR date_submitted < 2 ORDER BY date_submitted DESC;")

#AND Statement
dbGetQuery(mydb, "SELECT id_student, date_submitted FROM studentAssessment WHERE date_submitted > 550 AND id_student = 325750 ORDER BY date_submitted DESC;")

#LIKE
#%use% to find exact alike
dbGetQuery(mydb, "SELECT id_student, gender, region FROM studentInfo WHERE region LIKE '%Region%';")

#Begin with 'Region'
dbGetQuery(mydb, "SELECT id_student, gender, region FROM studentInfo WHERE region LIKE 'Region%';")

#End with 'Region'
dbGetQuery(mydb, "SELECT id_student, gender, region FROM studentInfo WHERE region LIKE '%Region';")

#'c' is the second letter
dbGetQuery(mydb, "SELECT id_student, gender, region FROM studentInfo WHERE region LIKE '_c%';")

#IN
#specific ketwords
dbGetQuery(mydb, "SELECT id_student, gender, region FROM studentInfo WHERE region IN ('Wales','Ireland');")

#EXERCISE 8
#Query one of your original toy data tables, for two different conditions. 

#ANSWER
#toy table toy2, IN
dbGetQuery(mydb, "SELECT n1, n2, n3 FROM toy2 WHERE n3 IN ('A','C');")

#toy table toy2, AND
dbGetQuery(mydb, "SELECT n1, n2 FROM toy2 WHERE n1 > 70 AND n2 < 20 ORDER BY n1 DESC;")

```

## Removing Duplicates
```{r}
dbGetQuery(mydb, "SELECT DISTINCT region FROM studentInfo;")

dbGetQuery(mydb, "SELECT DISTINCT region, gender FROM studentInfo;")

#EXERCISE 9
#Insert a duplicate row into one of your toy data tables. Then query the table without including duplicates.

#ANSWER
#insert a duplicate row into one of the toy tables
dbGetQuery(mydb,"INSERT INTO toy2 (n1, n2, n3) SELECT n1, n2, n3 FROM toy2 where n3 = 'B';")

#count rows
dbGetQuery(mydb, "SELECT COUNT(*) FROM toy2;")

#query without the dupicate row
dbGetQuery(mydb, "SELECT DISTINCT n3 FROM toy2;")

```

## Conditional Expressions (non-standard)
```{r}
dbGetQuery(mydb, "CREATE TABLE booltest (a INTEGER, b INTEGER);")
dbGetQuery(mydb, "INSERT INTO booltest VALUES (1, 0);")
dbGetQuery(mydb, "SELECT * FROM booltest;")

dbGetQuery(mydb,"SELECT
  CASE WHEN a THEN 'true' ELSE 'false' END as boolA,
  CASE WHEN b THEN 'true' ELSE 'false' END as boolB
  FROM booltest")

dbGetQuery(mydb,"SELECT
  CASE a WHEN 1 THEN 'true' ELSE 'false' END as boolA,
  CASE b WHEN 1 THEN 'true' ELSE 'false' END as boolB
  FROM booltest")
```

#Relationships (JOIN) - *Slide*
```{r}

#Create two tables with matches and join them

dbGetQuery(mydb, "CREATE TABLE left_table (id INTEGER, description TEXT);")
dbGetQuery(mydb, "CREATE TABLE right_table (id INTEGER, description TEXT);")

dbGetQuery(mydb, "INSERT INTO left_table VALUES ( 1, 'left 01');")
dbGetQuery(mydb, "INSERT INTO left_table VALUES ( 2, 'left 02');")
dbGetQuery(mydb, "INSERT INTO left_table VALUES ( 3, 'left 03');")
dbGetQuery(mydb, "INSERT INTO left_table VALUES ( 4, 'left 04');")
dbGetQuery(mydb, "INSERT INTO left_table VALUES ( 5, 'left 05');")
dbGetQuery(mydb, "INSERT INTO left_table VALUES ( 6, 'left 06');")
dbGetQuery(mydb, "INSERT INTO left_table VALUES ( 7, 'left 07');")
dbGetQuery(mydb, "INSERT INTO left_table VALUES ( 8, 'left 08');")
dbGetQuery(mydb, "INSERT INTO left_table VALUES ( 9, 'left 09');")

dbGetQuery(mydb, "INSERT INTO right_table VALUES ( 6, 'left 06');")
dbGetQuery(mydb, "INSERT INTO right_table VALUES ( 7, 'left 07');")
dbGetQuery(mydb, "INSERT INTO right_table VALUES ( 8, 'left 08');")
dbGetQuery(mydb, "INSERT INTO right_table VALUES ( 9, 'left 09');")
dbGetQuery(mydb, "INSERT INTO right_table VALUES ( 10, 'left 10');")
dbGetQuery(mydb, "INSERT INTO right_table VALUES ( 11, 'left 11');")
dbGetQuery(mydb, "INSERT INTO right_table VALUES ( 12, 'left 12');")
dbGetQuery(mydb, "INSERT INTO right_table VALUES ( 13, 'left 13');")
dbGetQuery(mydb, "INSERT INTO right_table VALUES ( 14, 'left 14');")

dbGetQuery(mydb, "SELECT * FROM left_table;")
dbGetQuery(mydb, "SELECT * FROM right_table;")

#join --> inner join, match two tables
dbGetQuery(mydb,"SELECT l.description AS left_table, r.description AS right_table 
           FROM left_table AS l 
           JOIN right_table AS r ON l.id = r.id")

#right join, right table = left table
dbGetQuery(mydb,"SELECT l.description AS left_table, r.description AS right_table 
           FROM left_table AS l 
           RIGHT JOIN right_table AS r ON l.id = r.id")

#left join, right_table matches the left one
dbGetQuery(mydb,"SELECT l.description AS left_table, r.description AS right_table 
           FROM left_table AS l 
           LEFT JOIN right_table AS r ON l.id = r.id")

#Union --> merge tables
dbGetQuery(mydb, "SELECT * FROM left_table
           UNION
           SELECT * FROM right_table;")

#what's difference between join and union?
#join --> match
#union --> merge


#EXERCISE 10
# Create a common id variable in your two toy data tables. Then join those tables so that your query returns all the values from one table and only those that match from the other. 

#Discussed with Professor Lang to create two new tables that I would know what's in them

#fourth attempt

dbGetQuery(mydb, "CREATE TABLE def5 (id INTEGER, num INTEGER, color TEXT);")
dbGetQuery(mydb, "CREATE TABLE def6 (id INTEGER, num INtEGER, color TEXT);")

#insert
dbGetQuery(mydb, "INSERT INTO def5 VALUES ( 1, 10, 'red');")
dbGetQuery(mydb, "INSERT INTO def5 VALUES ( 2, 20, 'blue');")
dbGetQuery(mydb, "INSERT INTO def5 VALUES ( 3, 30, 'green');")

dbGetQuery(mydb, "INSERT INTO def6 VALUES ( 3, 30, 'green');")
dbGetQuery(mydb, "INSERT INTO def6 VALUES ( 4, 40, 'violet');")
dbGetQuery(mydb, "INSERT INTO def6 VALUES ( 5, 50, 'yellow');")

#check tables
dbGetQuery(mydb, "SELECT * FROM def5;")
dbGetQuery(mydb, "SELECT * FROM def6;")

#LEFT join
dbGetQuery(mydb,"SELECT def5.num, def5.color, def6.num, def6.color 
           FROM def5
           LEFt JOIN def6 ON def5.id = def6.id")

#right join
dbGetQuery(mydb,"SELECT def5.num, def5.color, def6.num, def6.color 
           FROM def5
           RIGHT JOIN def6 ON def5.id = def6.id")


#check tables in the DB
#dbListTables(mydb)

#drop tables
#dbGetQuery(mydb,"DROP TABLE def5;")
#dbGetQuery(mydb,"DROP TABLE def6;")


###third attempt
## lines with single # worked
##df num1=1-3,num2=10-12,char=a
#df3 <- data.frame(id = seq(from = 1, to = 30, by = 1), num1 = c(rep(1:3,length.out=30)), num2 = c(rep(10:12,length.out=30)), char1=c(rep(letters[1],length.out=30)))

##id
##df3$id <- 1:nrow(df3)

##id = seq.int(nrow(df3))

##df num1=2, num2=11,char=a-b
#df4 <- data.frame(id = seq(from = 1, to = 30, by = 1), num1 = c(rep(2,30)), num2 = c(rep(11,30)), char1=c(rep(letters[1:2],length.out=30)))

##id
##df4$id <- 1:nrow(df4)


##call SQL package
#library(dplyr)

##left join, df3 < df4
##df34l <- left_join(df3, df4, by = c("id" = "id"))

#worked -
#dftest <- merge(df3,df4, all.x = TRUE)


##right join, df3 > df4
##df34r <- right_join(df3, df4, by = c("id" = "id"))

#worked -
#dftest2 <- merge(df3,df4, all.y = TRUE)

#sencond attempt
#r does not work, not same rows in either 1 or 2
#create two tables with three variables with ID

#df1 <- data.frame(id = seq(from = 1, to = 30, by = 1), num1 = c(rep(1:30)), num2 = c(rep_len(25:60,30)), char1=c(rep(letters[1:12],30))) 

#df2 <- data.frame(id = seq(from = 1, to = 30, by = 1), num1 = c(rep(5:35,length.out=30)), num2 = c(rep(15:40,length.out=30)), char1=c(rep(letters,length.out=30)))

#left join with df2 mathces df1
#library(dplyr)
#dfleft <- left_join(df1, df2)
#dfright <- right_join(df1, df2)


#first attemp, does not work with randomly generated tables
#create id variable in toynew1 and toy2

#add an ID Primary Key column
#dbGetQuery(mydb, "ALTER TABLE toynew1 ADD ID INTEGER AUTO_INCREMENT PRIMARY KEY") 
#dbGetQuery(mydb, "ALTER TABLE toy2 ADD ID INTEGER AUTO_INCREMENT PRIMARY KEY") 

#display
#dbGetQuery(mydb, "SELECT * FROM toynew1;")
#dbGetQuery(mydb, "SELECT * FROM toy2;")

#join the matching ones, use join
#dbGetQuery(mydb, "SELECT toynew1.n1, toynew1.n2, toynew1.n3, toynew1.ID, toy2.n1, toy2.n2, toy2.n3, toy2.ID 
#FROM toynew1
#LEFT JOIN toy2 ON 
#toynew1.ID = toy2.ID")
```
```{r}
#Now disconnect from your database
dbDisconnect(mydb)

#Then retunr to your AWS console and:

#1. Click on "Actions" and then "Stop"
#2. Do NOT make a snapshot
#3 Click on "Actions" again and click "Delete"
#4. Unclick "Make a final snapshot"
#5. Clicl "I acknowledge that upon instance deletion, automated backups, including system snapshots and point-in-time recovery, will no longer be available."
#6. Type "delete me" into the field

#Failure to follow these steps could result in charges to your credit card.


```

